Modal Verification Patterns for Systems Software

### 2025-09-13 paper.tex
+

Reasoning about the correctness of such code requires reasoning about these reachability relations and reference counts, relative to specific
stack frames.
% \paragraph{Buffering} Buffering is a technique used in variety of systems (filesystem page caching, write buffer in multi-processor architectures, 
% translation-lookaside-buffers for translated addresses etc.) to delay the costly operation (committing file pages, flushing buffered writes from write-cache 
% to memory etc.) making updates consistent.
% \paragraph{Consistency of Reads of Buffered Writes} Consistency of reads of any value potentially buffered depends on the semantics of 
% the operation making them \emph{observable}. A typical architecture such as Intel-x86 considers the \emph{flush} of buffered writes 
% \emph{non-deterministically} --- i.e., at any time at any number of buffered writes. On the other hand, a typical filesystem can follow a certain 
% policy to commit the in-memory file pages to the disk.   
\subsubsection{Weak Memory Models}
% I reworked this to be less general, and more about just weak memory, as thats all the survey section discusses.
Multi-core and multi-processor systems do not directly write every memory update back to memory (or even a coherent hardware cache)
individually, as this would be extremely slow. Instead, they effectively buffer reads and writes to and from memory
locally within each CPU core~\cite{owens2009better}\footnote{This is a layer below the better-known L1 cache, which is coherent across all CPUs in the system.}
so that operations can be partially reordered and batched to improve performance. For most code --- where the different CPUs are doing
independent work --- this is transparent to software. But fine-grained concurrent systems code which has multiple CPU cores
simultaneously coordinating access to shared memory must reason about these (effective) bufferings and reorderings, whose details vary
across CPU architectures.

Most of this coordination comes through reasoning about what \emph{will be true after local operations are published to memory},
or what \emph{will be true after remote operations become visible locally}~\cite{sieczkowski2015separation,dang2022compass,fsl}. Each of these flavors of contingent reasoning
depend on use of certain \emph{fence} or \emph{memory barrier} operations, which block execution of a CPU until either all incoming or outgoing
memory transactions are complete.
%\section{Resources in Systems Software}
%\label{sec:systemsoft}
%\subsection{Speculation in Consistency} \todo{speak about the weak-mem modalities in here.}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}
\section{Contingency Decomposition of a System}
In this section, we argue the position that \emph{modalities}  
should be a go-to approach when specifying and verifying low-level systems code.
We explain how the concept of a \emph{resource context} helps guide the design of new modalities
for verification of systems code, and
we justify our perspective by discussing prior systems that have used modalities for systems verification
successfully, arguing that they fit into the verification design pattern we articulate,
and explaining how this approach might apply to other systems verification challenges.

>>>>

Reasoning about the correctness of such code requires reasoning about these reachability relations and reference counts, relative to specific
stack frames.
% \paragraph{Buffering} Buffering is a technique used in variety of systems (filesystem page caching, write buffer in multi-processor architectures, 
% translation-lookaside-buffers for translated addresses etc.) to delay the costly operation (committing file pages, flushing buffered writes from write-cache 
% to memory etc.) making updates consistent.
% \paragraph{Consistency of Reads of Buffered Writes} Consistency of reads of any value potentially buffered depends on the semantics of 
% the operation making them \emph{observable}. A typical architecture such as Intel-x86 considers the \emph{flush} of buffered writes 
% \emph{non-deterministically} --- i.e., at any time at any number of buffered writes. On the other hand, a typical filesystem can follow a certain 
% policy to commit the in-memory file pages to the disk.   
\subsubsection{Weak Memory Models}
% I reworked this to be less general, and more about just weak memory, as thats all the survey section discusses.
Multicore and multiprocessor systems do not directly write every memory update back to memory (or even a coherent hardware cache)
individually, as this would be extremely slow. Instead, they effectively buffer reads and writes to and from memory
locally within each CPU core~\cite{owens2009better}\footnote{This is a layer below the better-known L1 cache, which is coherent across all CPUs in the system.}
so that operations can be partially reordered and batched to improve performance. For most code --- where the different CPUs are doing
independent work --- this is transparent to software. But fine-grained concurrent systems code which has multiple CPU cores
simultaneously coordinating access to shared memory must reason about these (effective) bufferings and reorderings, whose details vary
across CPU architectures.

Most of this coordination comes through reasoning about what \emph{will be true after local operations are published to memory},
or what \emph{will be true after remote operations become visible locally}~\cite{sieczkowski2015separation,dang2022compass,fsl}. Each of these flavors of contingent reasoning
depend on use of certain \emph{fence} or \emph{memory barrier} operations, which block execution of a CPU until either all incoming or outgoing
memory transactions are complete.
%\section{Resources in Systems Software}
%\label{sec:systemsoft}
%\subsection{Speculation in Consistency} \todo{speak about the weak-mem modalities in here.}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}
\subsubsection{Actor Model} An alternative to deal with the shared mmemory concurrency complication, i.e., data races, is using actors ~\cite{actormodel} which appears as a way of sequentializing the concurrent interleaved execution with some dependency on the \emph{knowledge} about the other actors.

\section{Contingency Decomposition of a System}
In this section, we argue the position that \emph{modalities}  
should be a go-to approach when specifying and verifying low-level systems code.
We explain how the concept of a \emph{resource context} helps guide the design of new modalities
for verification of systems code, and
we justify our perspective by discussing prior systems that have used modalities for systems verification
successfully, arguing that they fit into the verification design pattern we articulate,
and explaining how this approach might apply to other systems verification challenges.
